### 2015-05-11 swmpr_manu.tex
minor edits before submission

Occasionally, non-unique observations are present in the raw data.  These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data.  Erroneous duplicates with non-unique time stamps may also be present.  The \texttt{import\_local} function handles duplicate entries differently depending on the data type.  For water quality and nutrient data, duplicate time stamps are simply removed.  Nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.  Nutrient data with replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.  Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.  Only duplicate values at 15 minutes are averaged for weather data during import.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import local data for apaebmet
> 
> # this is an example path with the decompressed csv files
> path <- 'C:/my_path/'
> 
> # import, do not include file extension
> import_local(path, 'apadbwq') 
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

% attributes table
%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Attributes of a \texttt{swmpr} object that describe characteristics of the data.\label{tab:attributes}} 
\begin{center}
\begin{tabular}{lp{0.75in}p{3.25in}}
\hline\hline
\multicolumn{1}{l}{Attributes}&\multicolumn{1}{c}{Class}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{names}&character&Column names of the entire data set, inherited from the \texttt{data.frame} object class\tabularnewline
\texttt{row.names}&integer&Row names of the data set, inherited from the \texttt{data.frame} object class\tabularnewline
\texttt{class}&character&Class of the data object indicating \texttt{swmpr} and \texttt{data.frame}\tabularnewline
\texttt{station}&character&Station identifier used by \gls{NERRS} as a string with 7 or 8 characters\tabularnewline
\texttt{parameters}&character&Character vector of column names for data parameters, e.g., \texttt{`do\_mgl'}\tabularnewline
\texttt{qaqc\_cols}&logical&Indicates if \gls{QAQC} columns are present in the raw data\tabularnewline
\texttt{date\_rng}&POSIXct&Start and end dates for the raw data\tabularnewline
\texttt{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}\tabularnewline
\texttt{stamp\_class}&character&Class of the \texttt{datetimestamp} column, usually POSIXct unless data have been aggregated\tabularnewline
\hline
\end{tabular}\end{center}

@@

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the \gls{QAQC} flag that was assigned by \gls{CDMO} during initial processing.  The \gls{QAQC} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \gls{QAQC} flag (i.e., passed initial \gls{CDMO} checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

% table for organize functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} for individual functions or \\texttt{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?comb} for individual functions or \texttt{help.search(`organize', package = `SWMPr')} for all).\label{tab:organize}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{comb}&Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.\tabularnewline
\texttt{qaqc}&Remove \gls{QAQC} columns and remove data based on \gls{QAQC} flag values for a \texttt{swmpr} object.  Only applies if \gls{QAQC} columns are present. \tabularnewline
\texttt{qaqcchk}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different \gls{QAQC} flags used by \gls{CDMO}.  The output can be used to inform further processing.\tabularnewline
\texttt{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
\texttt{setstep}&Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.  The function is used in \texttt{comb} and can also be used with individual stations.\tabularnewline
\texttt{subset}&Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.\tabularnewline
\hline
\end{tabular}\end{center}

@@

Raw nutrient data obtained from the \gls{CDMO} will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call, including treatment of \texttt{NA}, values should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \gls{QAQC} columns are present.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get nutrient data
> data(apacpnut)
> dat <- apacpnut
> dat <- qaqc(dat)
> 
> # remove replicate nutrient data
> rem_reps(dat)
> 
> # use different function to aggregate replicates
> func <- function(x) max(x, na.rm = T)
> rem_reps(dat, FUN = func)
\end{verbatim}
\end{kframe}
\end{knitrout}

A subset method added to the existing generic \texttt{subset} function in R is available for \texttt{swmpr} objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator as a character string passed to the argument, such as \texttt{`>'} or \texttt{`<='}.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., `2007-01-01 06:30').  Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g. \texttt{`2012-11-31 12:00'}.  Finally, the function can be used to remove rows and columns that do not contain data.

@@

The \texttt{aggreswmp} function aggregates parameter data for a \texttt{swmpr} object by set units of time.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours by a predefined function, which defaults to the mean.  A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects created in the data.table package \cite{Dowle14}.  Additionally, the method of treating \texttt{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).

% table for analysis functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} for individual functions or \\texttt{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \texttt{?aggreswmp} for individual functions or \texttt{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{aggreswmp}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
\texttt{aggremetab}&Aggregate metabolism data from a \texttt{swmpr} object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of raw metabolism data.\tabularnewline
\texttt{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
\texttt{decomp}&Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.  This is a simple wrapper to \texttt{decompose} \cite{Kendall83}.  Decomposition of monthly or daily trends is possible.\tabularnewline
\texttt{decomp\_cj}&Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \texttt{decompTs} in the wq package \cite{Jassby14}.  Only monthly decomposition is possible.\tabularnewline
\texttt{hist}&Plot a histogram for a \texttt{swmpr} object.\tabularnewline
\texttt{lines}&Add lines to an existing plot created with \texttt{plot}.\tabularnewline
\texttt{map\_reserve}&Create a map of all stations in a reserve using the ggmap package.\tabularnewline
\texttt{na.approx}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
\texttt{plot}&Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.\tabularnewline
\texttt{plot\_metab}&Plot ecosystem metabolism estimates after running \texttt{ecometab} on a \texttt{swmpr} object.\tabularnewline
\texttt{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
\texttt{smoother}&Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \texttt{filter}.\tabularnewline
\hline
\end{tabular}\end{center}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import data, qaqc and subset
> data(apadbwq)
> dat <- qaqc(apadbwq)
> dat <- subset(dat, select = 'do_mgl', 
+   subset = c('2012-07-09 00:00', '2012-07-24 00:00')
+   )
> 
> # smooth
> dat_smooth <- smoother(dat, window = 50, params = 'do_mgl')
> 
> # plot raw and smoothed
> plot(dat)
> lines(dat_smooth, col = 'red', lwd = 2)
\end{verbatim}
\end{kframe}\begin{figure}[!h]

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get data
> data(apadbwq)
> swmp1 <- apadbwq
> 
> # subset for daily decomposition
> dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
> 
> # decomposition and plot
> test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
> plot(test)
\end{verbatim}
\end{kframe}\begin{figure}[!h]

@@

\caption[An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \texttt{decomp} function]{An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \texttt{decomp} function.}\label{fig:decomp_ex1}
\end{figure}

@@

An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package \cite{Cloern10,Jassby14}.  The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach works best for nutrient data that are typically obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data must first be aggregated on the monthly scale before decomposition.  Additionally, the time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.  For both, the random or events components can be considered anomalies that do not follow the trends in the remaining categories.  Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.  Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.  \Cref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.

@@

Water quality and weather observations from January 1995 to December 2014 for all \gls{NERRS} sites were obtained through a bulk data request using the zip downloads feature of \gls{CDMO}.  All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved locally as binary RData files (see \nameref{met_scripts}). This resulted in a single \texttt{swmpr} object for each parameter at each site.  All files were then uploaded to a remote server for online access.  An additional R script (see \nameref{met_scripts}) was executed that retrieved and processed water quality and weather data at each reserve to estimate metabolism.  Two water quality sites with the longest time series at each reserve were used.  Mean annual values at each site, organized by region, are shown in \Cref{fig:metab_plo}, whereas decadal comparisons are shown in \Cref{tab:metab_tab}.  All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with Caffrey \cite{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods. Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.  More importantly, use of the data retrieval, synthesis, and analysis functions to create the results illustrates the utility provided by the SWMPr package.

@@

%latex.default(metabs_tab[, !names(metabs_tab) %in% c("site",     "region")], file = "", caption = cap, cgroup = c("NEM\\textsuperscript{{\\it a}}",     "Pg", "Rt"), n.cgroup = c(2, 2, 2), colheads = rep(c("1995-2004",     "2005-2014"), 3), rowlabel = "Site", rowname = metabs_tab$site,     digits = 2, insert.bottom = foots, rgroup = unique(metabs_tab$region),     n.rgroup = nrgroups, label = "tab:metab_tab", size = "footnotesize",     where = "!tbp")%

@@

\multicolumn{1}{l}{\bfseries Site}&\multicolumn{2}{c}{\bfseries NEM\textsuperscript{{\it a}}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Pg}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Rt}\tabularnewline

@@

\footnotesize{\textsuperscript{{\it a}}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}\end{table}

>>>>

Occasionally, non-unique observations are present in the raw data.  These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data, or erroneous duplicates with non-unique time stamps.  The \texttt{import\_local} function handles duplicate entries differently depending on the data type.  For water quality and nutrient data, duplicate time stamps are simply removed.  Nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.  Nutrient data with replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.  Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.  Only duplicate values at 15 minutes are averaged for weather data during import.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import local data for apaebmet
> 
> # this is an example path with the decompressed csv files
> path <- 'C:/my_path'
> 
> # this is an example path for zipped csv files
> path <- 'C:/my_path.zip'
> 
> # import, do not include file extension
> import_local(path, 'apadbwq') 
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

% attributes table
%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Attributes of a \texttt{swmpr} object that describe characteristics of the data.\label{tab:attributes}} 
\begin{center}
\begin{tabular}{lp{0.75in}p{3.25in}}
\hline\hline
\multicolumn{1}{l}{Attributes}&\multicolumn{1}{c}{Class}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{names}&character&Column names of the entire data set, inherited from the \texttt{data.frame} object class.\tabularnewline
\texttt{row.names}&integer&Row names of the data set, inherited from the \texttt{data.frame} object class.\tabularnewline
\texttt{class}&character&Class of the data object indicating \texttt{swmpr} and \texttt{data.frame}.\tabularnewline
\texttt{station}&character&Station identifier used by \gls{NERRS} as a string with 7 or 8 characters.\tabularnewline
\texttt{parameters}&character&Character vector of column names for data parameters, e.g., \texttt{`do\_mgl'}, \texttt{`turb'}, etc.\tabularnewline
\texttt{qaqc\_cols}&logical&Indicates if \gls{QAQC} columns are present in the raw data.\tabularnewline
\texttt{date\_rng}&POSIXct&Start and end dates for the data.\tabularnewline
\texttt{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}.\tabularnewline
\texttt{stamp\_class}&character&Class of the \texttt{datetimestamp} column, usually POSIXct unless data have been aggregated.\tabularnewline
\hline
\end{tabular}\end{center}

@@

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified \gls{QAQC} flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \texttt{swmpr} object will have a corresponding \gls{QAQC} column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the \gls{QAQC} column range from -5 to 5 to indicate the \gls{QAQC} flag that was assigned by \gls{CDMO} during initial processing.  The \texttt{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \gls{QAQC} flag (i.e., passed initial \gls{CDMO} checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the \gls{QAQC} columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

% table for organize functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} for individual functions or \\texttt{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?comb} for individual functions or \texttt{help.search(`organize', package = `SWMPr')} for all).\label{tab:organize}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{comb}&Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.\tabularnewline
\texttt{qaqc}&Remove \gls{QAQC} columns and remove data based on \gls{QAQC} flag values for a \texttt{swmpr} object.  Only applies if \gls{QAQC} columns are present. \tabularnewline
\texttt{qaqcchk}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to each \gls{QAQC} flag used by \gls{CDMO}.  The output can be used to inform further processing.\tabularnewline
\texttt{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
\texttt{setstep}&Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.  The function is also used in \texttt{comb}.\tabularnewline
\texttt{subset}&Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.\tabularnewline
\hline
\end{tabular}\end{center}

@@

Raw nutrient data obtained from the \gls{CDMO} will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call, including treatment of \texttt{NA} values, should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \gls{QAQC} columns are present.

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get nutrient data, filter qaqc flags
> data(apacpnut)
> dat <- apacpnut
> dat <- qaqc(dat)
> 
> # remove replicate nutrient data
> rem_reps(dat)
> 
> # use different function to aggregate replicates
> func <- function(x) max(x, na.rm = T)
> rem_reps(dat, FUN = func)
\end{verbatim}
\end{kframe}
\end{knitrout}

A subset method added to the existing generic \texttt{subset} function in R is available for \texttt{swmpr} objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator as a character string passed to the \texttt{operator} argument, such as \texttt{`>'} or \texttt{`<='}.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., `2007-01-01 06:30').  Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g. \texttt{`2012-11-31 12:00'}.  Finally, the function can be used to remove rows and columns that do not contain data.

@@

The \texttt{aggreswmp} function aggregates parameter data for a \texttt{swmpr} object by set units of time.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours by a user-defined function, which defaults to the mean.  A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects created in the data.table package \cite{Dowle14}.  Additionally, the method of treating \texttt{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).

% table for analysis functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} for individual functions or \\texttt{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \texttt{?aggreswmp} for individual functions or \texttt{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{aggreswmp}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
\texttt{aggremetab}&Aggregate metabolism data from a \texttt{swmpr} object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
\texttt{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
\texttt{decomp}&Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.  This is a simple wrapper to \texttt{decompose} \cite{Kendall83}.  Decomposition of monthly or daily trends is possible.\tabularnewline
\texttt{decomp\_cj}&Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \texttt{decompTs} in the wq package \cite{Jassby14}.  Only monthly decomposition is possible.\tabularnewline
\texttt{hist}&Plot a histogram for a \texttt{swmpr} object.\tabularnewline
\texttt{lines}&Add lines to an existing plot created with \texttt{plot}.\tabularnewline
\texttt{map\_reserve}&Create a map of all stations in a reserve using the ggmap package \cite{Kahle13}.\tabularnewline
\texttt{na.approx}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
\texttt{plot}&Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.\tabularnewline
\texttt{plot\_metab}&Plot ecosystem metabolism estimates after running \texttt{ecometab} on a combined \texttt{swmpr} object.\tabularnewline
\texttt{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
\texttt{smoother}&Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \texttt{filter}.\tabularnewline
\hline
\end{tabular}\end{center}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import data, qaqc and subset
> data(apadbwq)
> dat <- qaqc(apadbwq)
> dat <- subset(dat, select = 'do_mgl', 
+   subset = c('2012-07-09 00:00', '2012-07-24 00:00')
+   )
> 
> # smooth using a window of 50 observations
> dat_smooth <- smoother(dat, window = 50, params = 'do_mgl')
> 
> # plot raw and smoothed
> plot(dat)
> lines(dat_smooth, col = 'red', lwd = 2)
\end{verbatim}
\end{kframe}\begin{figure}[!h]

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get data
> data(apadbwq)
> dat <- apadbwq
> 
> # subset for daily decomposition
> dat <- subset(dat, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
> 
> # daily decomposition of DO and plot
> dc_dat <- decomp(dat, param = 'do_mgl', frequency = 'daily')
> plot(dc_dat)
\end{verbatim}
\end{kframe}\begin{figure}[!h]

@@

\caption[An additive decomposition of dissolved oxygen into a trend, seasonal (daily), and random component using the \texttt{decomp} function]{An additive decomposition of dissolved oxygen into a trend, seasonal (daily), and random component using the \texttt{decomp} function.}\label{fig:decomp_ex1}
\end{figure}

@@

An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package \cite{Cloern10,Jassby14}.  The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach works best for nutrient data that are typically obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data must first be aggregated on the monthly scale before decomposition.  The time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.  For both, the random or events components can be considered anomalies that do not follow the trends in the remaining categories.  Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.  Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.  \Cref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.

@@

Water quality and weather observations from January 1995 to December 2014 for all \gls{NERRS} sites were obtained through a bulk data request using the zip downloads feature of \gls{CDMO}.  All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved locally as binary RData files (see \nameref{met_scripts}). This resulted in a single \texttt{swmpr} object for each parameter at each site.  All files were then uploaded to a remote server for online access.  An additional R script (see \nameref{met_scripts}) was executed that retrieved and processed water quality and weather data at each reserve to estimate metabolism.  Two water quality sites with the longest time series at each reserve were used.  Mean annual metabolism values at each site, organized by region, are shown in \Cref{fig:metab_plo}, whereas decadal comparisons are shown in \Cref{tab:metab_tab}.  All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with Caffrey \cite{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods. Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.  More importantly, use of the data retrieval, synthesis, and analysis functions to create the results illustrates the utility provided by the SWMPr package.

@@

%latex.default(metabs_tab[, !names(metabs_tab) %in% c("site",     "region")], file = "", caption = cap, cgroup = c("NEM\\textsuperscript{a}",     "Pg", "Rt"), n.cgroup = c(2, 2, 2), colheads = rep(c("1995-2004",     "2005-2014"), 3), rowlabel = "Site", rowname = metabs_tab$site,     digits = 2, insert.bottom = foots, rgroup = unique(metabs_tab$region),     n.rgroup = nrgroups, label = "tab:metab_tab", size = "footnotesize",     where = "!tbp")%

@@

\multicolumn{1}{l}{\bfseries Site}&\multicolumn{2}{c}{\bfseries NEM\textsuperscript{a}}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Pg}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries Rt}\tabularnewline

@@

\footnotesize{\textsuperscript{a}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}\end{table}

The second and third examples are two interactive web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \gls{SWMP} data.  The first web application evaluates trends in \gls{SWMP} data within and between sites using an interactive map (\cref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time.  Trends on the map at each station are plotted as circles that identify the direction and significance of the trend, such that larger points with darker colors indicate a strong trend.  The trend over time is blue for decreasing and red for increasing.  The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \texttt{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.  The output is identical to \cref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting. The data used for each application are similar to those used to estimate ecosystem metabolism described above and are available in the \nameref{supp_info}.  Both of the apps provide previously unavailable utilities to interact with \gls{SWMP} data using an intuitive online interface.  This format may be more appealing for individuals with limited experience using R, while simultaneously illustrating what is possible using functions from the package. The applications have been used extensively with over 300 hours noted in a single month.
